---
title: "Formatting custom geom bar unemployment plot"
title-block-banner: true
title-block-banner-color: azure
author: "Pablo Leon"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-location: left
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r markdown custom style, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 180, fig.width = 8, fig.height = 5)
library(tidyverse) 
```

## Turning out of the box ggplot2 chart into a custom plot

This is a Quarto document describing the steps required to format a standard ggplot2 bar plot adding some theme options and annotation bubble with specific highlighting and color settings.

## 1.Load unemployment data

```{r Import unemployment data, echo=TRUE}
# 1. Import previous files.
library(here)
library(tidyverse)

# 1. Import unemployment indicator data 
unemp_data <- read.table(here("data_cleansed", "EU_UNEMP_CLEAN_une_rt_a_LONG.csv"),
                         header =TRUE, sep =',',stringsAsFactors =TRUE)
head(unemp_data)
```

## 2.Subset initial dataset to keep just Greece data

You can add options to executable code like this

```{r}
# 2. Subset data for Greece
str(unemp_data)
head(unemp_data)

unemp_greece <- unemp_data %>% 
  filter(country %in% c("greece")) %>%  
  select(-c(X))
unemp_greece
```

The `echo: false` option disables the printing of code (only output is displayed).

## 3.Create two flags to highlight years in plot

We will create two specific flags, one of them to display in a different color the year with highest unemployment rate, and a second flag to highlight the most recent year of data.

```{r Create Flag variables, echo=TRUE}
# 3. Create two flags
# 3.1. Flag for highest unemployment year
str(unemp_greece)

unemp_greece_max <- unemp_greece %>% 
  mutate(
    Date = as.Date(date),
    unemp_round = round(unemp_rate,0),
    max_unemp = max(unemp_rate, na.rm = TRUE),
    Max_unemp_flag = ifelse(unemp_rate == max_unemp,TRUE,FALSE) 
  ) %>% 
  select(date,Date,country,unemp_rate,max_unemp,Max_unemp_flag)

str(unemp_greece_max)

unemp_greece_latest <- unemp_greece_max %>% 
  mutate(
      Latest_date = max(Date),
      Max_date_flag = ifelse(Date == Latest_date,TRUE,FALSE))
```

Check first rows including now the new flag variables created

```{r Check new flags, echo=TRUE}
head(unemp_greece_latest)
```

```{r display new columns created, echo=TRUE}
names(unemp_greece_latest)
```

## 4. Plot basic bar plot using ggplot 

This is going to be the starting point of formatting a standard ggplot2 bar plot. It only contains the basic element and a single geom() to plot a simple bar plot.

```{r Basic ggplot2 bar plot, echo=TRUE}
# Create a basic bar plot
str(unemp_greece)
Plot01 <- unemp_greece %>% 
  ggplot(aes(x = date, y = unemp_rate)) +
  geom_col(fill = "#BAD1D6")

Plot02 <- unemp_greece %>% 
  ggplot(aes(x = date, y = unemp_rate)) +
  geom_col(fill = "#BAD1D6") +
labs(title = "Unemployment in Greece.2003-2023 period",
     caption = "Note: Year 2023  latest available data. Source:EUROSTAT https://ec.europa.eu/eurostat/") +
  theme_classic() 

Plot02
ggsave("plots_output/10_Greece_unemployment_value.png", width = 6.38, height = 5.80)

```

