---
title: "Recent Spain population trends"
author: "PLR"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup include requried libraries, include=FALSE}
pacman::p_load(readxl,here,dplyr,janitor,ggplot2,gt,lubridate,plyr,tidyr,stats,data.table)
```

## Latest date this report was produced

Today's date is **`r format(today(),"%d %B %Y")`**. This report was published on the week starting on  **`r format(today()-2,"%d %B %Y")`**. 


## 1. Load Spain population data

We load latest Spanish Population Figures released by the Spanish National Institute: <https://www.ine.es/consul/serie.do?d=true&s=ECP320>

In the third tab from Excel file **"INE total and foreign population figures Spain.xlsx"** we load the table describing Total, foreign and nationals population figures in Spain, based on Spanish nationality for the 2004 2025 period.

```{r Load spanish population data, echo=FALSE, message=FALSE, warning = FALSE}
list.files (path = "./data_demography" ,pattern = "xlsx$")
excel_sheets("./data_demography/INE total and foreign population figures Spain.xlsx")
```


```{r read Excek into R, echo=TRUE, message=FALSE, warning = FALSE}
population_data <- read_excel(
  "data_demography/INE total and foreign population figures Spain.xlsx",
  sheet = "INE_Total_foreign_population", skip = 2, n_max = 22,
  col_types = c("guess", "numeric", "numeric", "guess","guess","guess","guess","guess")) %>% 
  clean_names() %>% 
       select (date = todas_las_edades, total_population = total,
                       foreign_population, percent_foreign_population = percent_foreign_nationals_total_population,
                      total_population_YoY_N = total_yo_y_n, total_population_YoY_perc = total_yo_y_percent,
                       foreign_population_YoY_N = foreign_nationals_yo_y_n, foreign_population_YoY_perc= foreign_total_yo_y_percent)

head(population_data)
names(population_data)
```


## 2. Exploratory charts Total, Spanish nationals and foreign population in Spain

From the newly on boarded data, we create a new Year variable from initial date column. As we only want to display Year values.

```{r cleansing_inpported data, echo=TRUE, message=FALSE, warning = FALSE}
INE_population_subset <- population_data %>%
                         select(date,total_population,foreign_population) %>% 
                         mutate(Year = substring(date, 15, 25)) 
INE_population_subset
```
```{r creating_first_calculations, echo=TRUE, message=FALSE, warning = FALSE}
INE_calc_fields <- INE_population_subset %>%                          
                         select(Year,total_population,foreign_population) %>% 
                         mutate(Spanish_nationals = total_population - foreign_population) %>% 
                         arrange(Year)
INE_calc_fields
```

These charts below describe the evolution of Total, foreign and spanish nationals population for the 2005-2025 period:

- Spain total population
```{r Spain total population bar plot, echo=TRUE,warning=FALSE,message=FALSE}
options(scipen=999)

Spanish_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = total_population)) +
  geom_bar(stat = "identity", fill = "darkolivegreen2") +
  labs(title = "Spain total poulation. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +
  theme(axis.text.x = element_text(angle = + 90, hjust = 0.5, vjust = 0.5)) +
  geom_text(aes(label = total_population),size = 1.8,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,55000000), expand = FALSE )

Spanish_population_plot
```

Then this is the foreign population in Spain for the same time period.

```{r Spain foreign population bar plot, echo=TRUE,warning=FALSE,message=FALSE}
options(scipen=999)

forign_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = foreign_population)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(title = "Foreign population in Spain. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +
  theme(axis.text.x = element_text(angle = + 90, hjust = 0.5, vjust = 0.5)) +
  geom_text(aes(label = foreign_population),size = 1.8,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,7500000), expand = FALSE )

forign_population_plot
```


This last plot displays national population in Spain for the 2005-2025 period

```{r Spain national population bar plot, echo=TRUE, warning=FALSE,message=FALSE}
options(scipen=999)

national_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = Spanish_nationals)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Spanish nationals population in Spain. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +
  theme(axis.text.x = element_text(angle = + 90, hjust = 0.5, vjust = 0.5)) +
  geom_text(aes(label = Spanish_nationals),size = 1.8,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,46000000), expand = FALSE )

national_population_plot
```


## 3.  Spain population figures Overveiw details 

In this section we highlight a couple of trends observed in both Total and foreign population in Spain from 2004 until 2025 period.

```{r Population figures overview, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Total population figures overview

Max_total_population <- INE_calc_fields  %>% select (Year,total_population) %>% filter(Year == max(INE_calc_fields$Year))
Max_total_population


Min_total_population <- INE_calc_fields  %>% select (Year,total_population) %>% filter(Year == min(INE_calc_fields$Year))
Min_total_population

# Foreign population figures overview
foreign_pop_first_year <- INE_calc_fields  %>% select (Year,foreign_population) %>% filter(Year == 2005)

foreign_pop_last_year <- INE_calc_fields  %>% select (Year,foreign_population) %>% filter(Year == 2025)
```

- As of 1st of January **`r (Max_total_population$Year)`**., the year for the latest available population figures, Total population in Spain was  **`r prettyNum(Max_total_population$total_population, big.mark=",")`**. Up by 458,253 from previous year.

- In contrast, on the year **`r (Min_total_population$Year)`**, the first year on this series, total population in Spain was **`r prettyNum(Min_total_population$total_population, big.mark=",")`**. 

- In terms of foreign population, on the following year **`r (foreign_pop_first_year$Year)`**.there was a foreign population of **`r prettyNum(foreign_pop_first_year$foreign_population, big.mark=",")`** 

- At the end of the series on 1st January **`r (foreign_pop_last_year$Year)`**, latest foreign population figures in Spain was **`r prettyNum(foreign_pop_last_year$foreign_population, big.mark=",")`**. We will describe these population changes in absolute figures and percent change in the next section below.

### 3.1 Share of foreign population over total population in Spain

Next, we describe the foreign population percent share of the total population in Spain for the 2004-2025 period.


## 4. Populating report using helper functions

In this section we will use a set of helper functions to create initial data sets for selected countries. And also to format each figure used in the Markdwon report. 


```{r helper functions,  include=TRUE, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}
# I need to place all my Functions in this chunck to be used in the report !!! 
# So I can render the Markdown report !!!
# FUNCTION 01 02 - Import indicators
Import_eurostat_indicators <- function(tab_name,choose_directory = NULL, selected_countries,indicator = NULL){

  # une_rt_a (Unemployment by sex and age - annual data). Time 23/23 (2003-2025)
  if (indicator == "unemp"){
  # 1.1 arange original input data in Long format  
  unemp_raw <- read_excel(file.path(here::here(), "data","une_rt_a__custom_14324113_page_spreadsheet.xlsx"),
                              sheet = tab_name, col_names = TRUE, na = ":", skip = 8,n_max = 23) %>% 
              filter(!is.na(France)) %>%  # France has the highest number of populated rows only 1 NA
              pivot_longer(!Date, names_to = "Countries", values_to = "metric_value") 
  unem_long <- unemp_raw %>% mutate(metric = "unemployment_rate", units = "percentage") %>% 
                             select(date = Date,country = Countries,metric_value, metric, units) %>% 
                             mutate(metric_value = as.numeric(metric_value)) %>% # to compute calculations metric_Value must be numeric
                         filter(country %in% c(selected_countries))   #  filter initial data by selection of countries
  # 1.2 New variable - lagged values (1year ago, 2 years ago, 5 years ago, grouped by country)
  # date_1y_ago, value_1y_ago, date_5y_ago, value_5y_ago
  unem_long_lags <- unem_long %>% 
                    arrange(country,date) %>% 
                    group_by(country) %>% 
                    mutate(
                      date_1y_ago = lag(date,1),
                      value_1y_ago = lag(metric_value,1),
                      date_2y_ago = lag(date,2),
                      value_2y_ago = lag(metric_value,2),
                      date_5y_ago = lag(date,5),
                      value_5y_ago = lag(metric_value,5)
                      ) %>% 
                    ungroup()
  # 1.3 Add new set of columns to display min and max values BY COUNTRY
  unemp_long_min_max<- unem_long_lags %>%
                   select(country,date,metric_value,metric,units) %>%
    group_by(country) %>% 
  mutate(
          min_value_country = min(metric_value, na.rm = TRUE),
          max_value_country = max(metric_value, na.rm = TRUE)
          ) %>% 
    ungroup()
  # 1.5 Finally include min and max values entire unemp dataset
  unempl_all <- unemp_long_min_max %>% 
    mutate(
      min_value_indic = min(metric_value, na.rm = TRUE),
      max_value_indic = max(metric_value, na.rm = TRUE)
    )              
  
  # 1.6 Ensure final output from function is a data.frame() object
  unemp_long_dataframe <- data.frame(unempl_all)
  
  # 1.7 Write dataframe to "data_cleansed_folder"
  filename <- "country_sel_unemp_long_dataframe.csv"
  output_file <- file.path("data_cleansed",filename)   # Using file.path() to build relative path to data_cleansed sub-folder. works on Windows, Linux, and macOS.
  write.csv(unemp_long_dataframe,file = output_file,row.names = FALSE)
  cat("File saved as:", output_file, "\n") # Write  message on Console everytime the output file is written to .csv and saved to "data_cleansed" sub-folder
  
  # Return final selection of countries unemployment indicator values    
  return(unemp_long_dataframe)
  
  } else if (indicator == "part_time_persons"){
  # Downloaded table data: lfsi_pt_a (Part-time employment and temporary contracts-annual data)
  # Return final selection of countries temporary employment figures
  part_time_emp_raw <- read_excel(file.path(here::here(),"data","lfsi_pt_a__custom_14828862_page_spreadsheet.xlsx"),
                             sheet = tab_name, col_names = TRUE, na = ":", skip = 10, n_max = 22) %>% 
    filter(!is.na(France)) %>%  # France has the highest number of populated rows only 1 NA
    pivot_longer(!Date, names_to = "Countries", values_to = "metric_value") 
  
  part_time_long <- part_time_emp_raw %>% mutate(metric = "per_persons_working_pat_time", units = "percentage") %>% 
               select(date = Date,country = Countries,metric_value, metric, units) %>% 
    filter(country %in% c(selected_countries))   #  filter initial data by selection of countries
  
  # Return final selection of countries temporary employment indicator values  
part_time_long_lags <- part_time_long %>% 
    arrange(country,date) %>% 
    group_by(country) %>% 
    mutate(
      date_1y_ago = lag(date,1),
      value_1y_ago = lag(metric_value,1),
      date_2y_ago = lag(date,2),
      value_2y_ago = lag(metric_value,2),
      date_5y_ago = lag(date,5),
      value_5y_ago = lag(metric_value,5)
    ) %>% 
    ungroup()

part_time_long_min_max<- part_time_long_lags %>%
  select(country,date,metric_value,metric,units) %>%
  group_by(country) %>% 
  mutate(
    min_value_country = min(metric_value, na.rm = TRUE),
    max_value_country = max(metric_value, na.rm = TRUE)
  ) %>% 
  ungroup()
# 1.5 Finally include min and max values entire unemp dataset
part_time_all <- part_time_long_min_max %>% 
  mutate(
    min_value_indic = min(metric_value, na.rm = TRUE),
    max_value_indic = max(metric_value, na.rm = TRUE)
  )              
# 1.6 Ensure final temp_emp dataframe output from function is a data.frame() object
temp_emp_long_dataframe <- data.frame(part_time_all)

return(temp_emp_long_dataframe)
    }
  
}
# Parameters (tab_name = "Sheet 1", selcted_countries = c("country1","country2","country3"), indicator ="unemp/part_time_persons")
Import_eurostat_indicators(tab_name = "Sheet 1", selected_countries = c('Bulgaria','Estonia','Ireland'),indicator = "unemp")

```
```{r helper function 02, echo=TRUE, warning = FALSE, message = FALSE}

# 02 Function - declares path to data folder where formatted input data is saved  

# FUNCTION 02 - Declare filepath to "data_cleansed" folder
data_filepath  <- function(tab_name = NULL,choose_directory = NULL, own_directory = NULL){
  
  if(choose_directory == "data_folder") {
    data_folder_path = file.path(here::here(), "data") 
    if (dir.exists(data_folder_path)) {return(data_folder_path)}
  } else if (choose_directory == "data_cleansed") {
    data_cleansed_path = file.path(here::here(),"data_cleansed")
    if (dir.exists(data_cleansed_path)) {return(data_cleansed_path)}
  } else { stop ("please provide your own directory")}
  # Include details about user directory
  if (dir.exists(own_directory)){return(own_directory)}
  else{stop("Please ensure you provide your own_directory",own_directory)}
  
}

# Use function
data_filepath(choose_directory = "data_cleansed")

```



```{r Input previouos into Markdown report, echo=TRUE,warning=FALSE,message=FALSE}
data_filepath(choose_directory = "data_cleansed")

dataset_sel_countries <- fread("data_cleansed/country_sel_unemp_long_dataframe.csv")
dataset_sel_countries
```


Now we use function from previous r code chunch called "Import_eurostat_indicators()" to subset unemployment indicator data for a selection of countries ("Bulgaria","Estonia" and "Ireland") 

Besides, in this section we can see how to populate text using "fmt_markdown_figures" function below to apply specific format types (numeric values including thousand separators, and percentage values displaying the "%" sign) when describing figures in the rendered markdown text output file.

```{r Function subset indicators for selected countries, echo=FALSE}
Import_eurostat_indicators(tab_name = "Sheet 1", selected_countries = c('Bulgaria','Estonia','Ireland'),indicator = "unemp")
```

- Now I include a new function in the section below to display unemployment figures using inline R code with specific formats. 

- This formats are defined in the function below called "fmt_markdown_figures" and it allows us to add thousands separator when the indicator is an integer number and % sign when it displays percentages such as in the unemployment rate example below.

## 5. Unemployment rate in 2009 

- This section below includes formatted figures using auxiliary function "fmt_markdown_figures()" from helper_functions.R script:

```{r formatting function, echo=FALSE}

fmt_markdown_figures<- function(mydataset,countryname,datevalue,column,format = NULL){
  row <- mydataset %>% filter(country == countryname) 
  row_date <- row %>%  filter(date == datevalue)
  value <- row_date %>% pull({{column}})
  units <- row_date %>% pull(units)
  # safe checks
  if (length(value)==0){return(NA)}
  if (is.null(format)){format <- units}
  
  # 1. Section to start applying required formats using format parameter
  # 1.1 numeric format
  if (format == "numeric"){
      if(is.na(value)){             # Accounting for missing values When there are NA values in original input data
      return(NA_character_)
    }   else if (!is.na(value)){   # Ensure value is not missing so it is valied 
  value_num <- as.numeric(value) # I need to ensure is numeric to multiply it by 100
  return((prettyNum(value_num*10000,big.mark = ","))) # Just testing multiplying it by 1000 to see the big mark displayed
  } 
  # 1.2 percentage format  
  } else if (format == "percentage"){
  return(paste0(round(value,1),"%"))  # to be built
  }
  # Default value if format is not provided
  return(as.character(prettyNum(value,big.mark = ",")))
}
```


In 2009 unemployment rate for Bulgaria was `r fmt_markdown_figures(mydataset = dataset_sel_countries, countryname = "Bulgaria",datevalue = 2010, column = "metric_value",format ="percentage")` compared to 
 `r fmt_markdown_figures(mydataset = dataset_sel_countries, countryname = "Bulgaria",datevalue = 2009, column = "metric_value",format ="percentage")` one year before in 2009. 
 
 This is an increase of 