---
title: "Recent Spain population trends"
author: "PLR"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
# Load required libraries
library(pacman)
pacman::p_load(readxl,here,dplyr,janitor,ggplot2,gt,lubridate,plyr,tidyr,stats,data.table)
# Apply consistent height and width ggplot graph across document
knitr::opts_chunk$set(echo = TRUE, fig.width = 20, fig.height = 10)
```

## Latest date this report was produced

Today's date is **`r format(today(),"%d %B %Y")`**. This report was published on the week starting on  **`r format(today()-2,"%d %B %Y")`**. 


## 1. Load Spain population data

We load latest Spanish Population Figures released by the Spanish National Institute: <https://www.ine.es/consul/serie.do?d=true&s=ECP320>

In the third tab from Excel file **"INE total and foreign population figures Spain.xlsx"** we load the table describing Total, foreign and nationals population figures in Spain, based on Spanish nationality for the 2004 2025 period.

```{r Load spanish population data, echo=FALSE, message=FALSE, warning = FALSE}
list.files (path = "./data_demography" ,pattern = "xlsx$")
excel_sheets("./data_demography/INE total and foreign population figures Spain.xlsx")
```


```{r read Excek into R, echo=TRUE, message=FALSE, warning = FALSE}
population_data <- read_excel(
  "data_demography/INE total and foreign population figures Spain.xlsx",
  sheet = "INE_Total_foreign_population", skip = 2, n_max = 22,
  col_types = c("guess", "numeric", "numeric", "guess","guess","guess","guess","guess")) %>% 
  clean_names() %>% 
       select (date = todas_las_edades, total_population = total,
                       foreign_population, percent_foreign_population = percent_foreign_nationals_total_population,
                      total_population_YoY_N = total_yo_y_n, total_population_YoY_perc = total_yo_y_percent,
                       foreign_population_YoY_N = foreign_nationals_yo_y_n, foreign_population_YoY_perc= foreign_total_yo_y_percent)

head(population_data)
names(population_data)
```


# 2. Spanish total population change over time

This sections described total population change in Spain from 


```{r Total population, echo=TRUE, message=FALSE, warning = FALSE}

population_change <- population_data %>% select(date,total_population,foreign_nationals_population = foreign_population,percent_foreign_population,total_population_YoY_perc,foreign_population_YoY_perc)
population_change

```
```{r}
library(lubridate)
population_change_date_creation <- population_change %>% 
  mutate(date_day = substr(date,1,1),
         date_month = substr(date,3,13),
         date_year =  substr(date,15,18))

population_change_fmt <-population_change_date_creation %>% 
mutate(date_month_eng = gsub("de enero de","january",date_month)) %>% 
select(date,total_population,foreign_nationals_population,percent_foreign_population,total_population_YoY_perc,foreign_population_YoY_perc,date_day,date_month_eng,date_year) %>% 
mutate(date_to_fmt = paste0(date_day," ",date_month_eng," ",date_year)) 


population_change_fmt_date <- population_change_fmt %>% 
  mutate(date_fmt = dmy(date_to_fmt)) %>% 
  select(date_fmt,total_population,foreign_nationals_population,percent_foreign_population,total_population_YoY_perc_change = total_population_YoY_perc,foreign_population_YoY_perc_change = foreign_population_YoY_perc) %>% 
  arrange(date_fmt)
population_change_fmt_date

min_date <- min(population_change_fmt_date$date_fmt)
max_date <- max(population_change_fmt_date$date_fmt)
# (gsub("de enero","january", date),1,6)),
```
## 2.1 Spain total population

This first chart displays total population in Spain for 2004-2025 time period. Total resident population in Spain, regardless of their nationality 

```{r format_function_markdown_values, echo=TRUE,warning=FALSE,message=FALSE, results='hide'}
fmt_markdown_figures<- function(mydataset,countryname,datevalue,column,format = NULL){
  row <- mydataset %>% filter(country == countryname) 
  row_date <- row %>%  filter(date == datevalue)
  value <- row_date %>% pull({{column}})
  units <- row_date %>% pull(units)
  # safe checks
  if (length(value)==0){return(NA)}
  if (is.null(format)){format <- units}
## 1. Section to start applying required formats using format parameter
  if (format == "numeric"){
      if(is.na(value)){             # Accounting for missing values When there are NA values in original input data
      return(NA_character_)
    }   else if (!is.na(value)){   # Ensure value is not missing so it is valied 
  value_num <- as.numeric(value) # I need to ensure is numeric to multiply it by 100
  return((prettyNum(value_num*10000,big.mark = ","))) # Just testing multiplying it by 1000 to see the big mark displayed
  } 
    
  } else if (format == "percentage"){
  return(paste0(round(value,1),"%"))  # to be built
  }
  # Default value if format is not provided
  return(as.character(prettyNum(value,big.mark = ",")))
}
```

```{r Spain total population, echo=TRUE,message=FALSE, warning =FALSE}
Total_pop_spain <- population_change_fmt_date %>% select(date_fmt,total_population) %>% 
  ggplot(aes(date_fmt,total_population)) +
  geom_line(aes(colour = "sienna3")) +
  geom_point(fill = "sienna3") +
  labs(title = "Spain total poulation. 2004-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  scale_y_continuous(breaks = seq(40000000, 50000000, by = 200000)) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year") +
  theme_light() +
   theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = + 45, hjust = 0.5, vjust = 0.5),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) 
Total_pop_spain


```




## 3. Exploratory charts Total, Spanish nationals and foreign population in Spain

From the newly on boarded data, we create a new Year variable from initial date column. As we only want to display Year values.

```{r cleansing_inpported data, echo=TRUE, message=FALSE, warning = FALSE}
INE_population_subset <- population_data %>%
                         select(date,total_population,foreign_population) %>% 
                         mutate(Year = substring(date, 15, 25)) 
INE_population_subset
```
```{r creating_first_calculations, echo=TRUE, message=FALSE, warning = FALSE}
INE_calc_fields <- INE_population_subset %>%                          
                         select(Year,total_population,foreign_population) %>% 
                         mutate(Spanish_nationals = total_population - foreign_population) %>% 
                         arrange(Year)
INE_calc_fields
```

These charts below describe the evolution of Total, foreign and spanish nationals population for the 2005-2025 period:

- Spain total population
```{r Spain total population bar plot, echo=TRUE,warning=FALSE,message=FALSE}
options(scipen=999)

Spanish_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = total_population)) +
  geom_bar(stat = "identity", fill = "darkolivegreen2") +
  labs(title = "Spain total poulation. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +
  theme(axis.text.x = element_text(angle = + 90, hjust = 0.5, vjust = 0.5)) +
  geom_text(aes(label = total_population),size = 2.8,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,55000000), expand = FALSE )

Spanish_population_plot
```

Then this is the foreign population in Spain for the same time period.

```{r Spain foreign population bar plot, echo=TRUE,warning=FALSE,message=FALSE}
options(scipen=999)

forign_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = foreign_population)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(title = "Foreign population in Spain. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +
  theme(axis.text.x = element_text(angle = + 90, hjust = 0.5, vjust = 0.5)) +
  geom_text(aes(label = foreign_population),size = 1.8,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,7500000), expand = FALSE )

forign_population_plot
```


This last plot displays national population in Spain for the 2005-2025 period.

```{r Spain national population bar plot, echo=TRUE, warning=FALSE,message=FALSE}
options(scipen=999)

national_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = Spanish_nationals)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Spanish nationals population in Spain. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +
  theme(axis.text.x = element_text(angle = + 90, hjust = 0.5, vjust = 0.5)) +
  geom_text(aes(label = format(Spanish_nationals,big.mark = ",",scientific = FALSE))
            ,size = 4.2,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,46000000), expand = FALSE )

national_population_plot
```


## 3.  Spain population figures Overview details 

In this section we highlight a couple of trends observed in both Total and foreign population in Spain from 2004 until 2025 period.

```{r Population figures overview, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Total population figures overview

Max_total_population <- INE_calc_fields  %>% select (Year,total_population) %>% filter(Year == max(INE_calc_fields$Year))
Max_total_population


Min_total_population <- INE_calc_fields  %>% select (Year,total_population) %>% filter(Year == min(INE_calc_fields$Year))
Min_total_population

# Foreign population figures overview
foreign_pop_first_year <- INE_calc_fields  %>% select (Year,foreign_population) %>% filter(Year == 2005)

foreign_pop_last_year <- INE_calc_fields  %>% select (Year,foreign_population) %>% filter(Year == 2025)
```

- As of 1st of January **`r (Max_total_population$Year)`**., the year for the latest available population figures, Total population in Spain was  **`r prettyNum(Max_total_population$total_population, big.mark=",")`**. Up by 458,253 from previous year.

- In contrast, on the year **`r (Min_total_population$Year)`**, the first year on this series, total population in Spain was **`r prettyNum(Min_total_population$total_population, big.mark=",")`**. 

- In terms of foreign population, on the following year **`r (foreign_pop_first_year$Year)`**.there was a foreign population of **`r prettyNum(foreign_pop_first_year$foreign_population, big.mark=",")`** 

- At the end of the series on 1st January **`r (foreign_pop_last_year$Year)`**, latest foreign population figures in Spain was **`r prettyNum(foreign_pop_last_year$foreign_population, big.mark=",")`**. We will describe these population changes in absolute figures and percent change in the next section below.

### 3.1 Share of foreign population over total population in Spain

Next, we describe the foreign population percent share of the total population in Spain for the 2004-2025 period.
