---
title: "Recent Spain population trends"
author: "PLR"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(here)
library(dplyr)
library(janitor)
library(ggplot2)
library(gt)
library(lubridate)
```

## Latest date this report was produced

Today's date is **`r format(today(),"%d %B %Y")`**. This report was published on the week starting on  **`r format(today()-2,"%d %B %Y")`**. 


## 1. Load Spain population data

We load latest Spanish Population Figures released by the Spanish National Institute: <https://www.ine.es/consul/serie.do?d=true&s=ECP320>

In the third tab from Excel file **"INE total and foreign population figures Spain.xlsx"** we load the table describing Total, foreign and nationals population figures in Spain, based on Spanish nationality for the 2004 2025 period.

```{r Load spanish population data, echo=FALSE}
list.files (path = "./data_demography" ,pattern = "xlsx$")
excel_sheets("./data_demography/INE total and foreign population figures Spain.xlsx")
```


```{r read Excek into R, echo=TRUE}
population_data <- read_excel(
  here("data_demography", "INE total and foreign population figures Spain.xlsx"), 
  sheet = "INE_Total_foreign_population", skip =2, n_max = 22,
  col_types = c("guess", "numeric", "numeric", "guess","guess","guess","guess","guess")) %>% 
  clean_names() %>% 
               select (date = todas_las_edades, total_population = total,
                       foreign_population, percent_foreign_population = percent_foreign_nationals_total_population,
                       total_population_YoY_N = total_yo_y_n, total_population_YoY_perc = total_yo_y_percent,
                       foreign_population_YoY_N = foreign_nationals_yo_y_n, foreign_population_YoY_perc= foreign_total_yo_y_percent)
population_data

head(population_data)
names(population_data)
```


## 2. Cleanse Spain populations figures data

From the newly on boarded data, we create a new Year variable from initial date column. As we only want to display Year values.

```{r cleansing_inpported data, echo=TRUE}
INE_population_subset <- population_data %>%
                         select(date,total_population,foreign_population) %>% 
                         mutate(Year = substring(date, 15, 25)) 
INE_population_subset
```
```{r creating_first_calculations, echo=TRUE}
INE_calc_fields <- INE_population_subset %>%                          
                         select(Year,total_population,foreign_population) %>% 
                         mutate(Spanish_nationals = total_population - foreign_population) %>% 
                         arrange(Year)
INE_calc_fields
```

These charts below describe the evolution of Total, foreign and spanish nationals population for the 2005-2025 period:

- Spain total population
```{r Spain total population bar plot, echo=TRUE}
options(scipen=999)

Spanish_population_plot <- INE_calc_fields %>% 
                            ggplot(aes(x= Year, y = total_population)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Total population in Spain. 2005-2025 period",
       substile = "Source: INE Spanish Office for National Statistics") +
  theme_light() +

  geom_text(aes(label = total_population),size = 1.8,position = position_dodge(width = 0.2),vjust = -0.30,hjust = 0.50) +
  coord_cartesian( ylim=c(0,55000000), expand = FALSE )

Spanish_population_plot
```
Then this is the foreign population in Spain for the same time period.
```{r Spain total population bar plot, echo=TRUE}
INE_calc_fields <- INE_population_subset %>%                          
                         select(Year,total_population,foreign_population) %>% 
                         mutate(Spanish_nationals = total_population - foreign_population) %>% 
                         arrange(Year)
INE_calc_fields
```
This last plot displays national population in Spain for the 2005-2025 period
```{r Spain total population bar plot, echo=TRUE}
INE_calc_fields <- INE_population_subset %>%                          
                         select(Year,total_population,foreign_population) %>% 
                         mutate(Spanish_nationals = total_population - foreign_population) %>% 
                         arrange(Year)
INE_calc_fields
```


## 3.  Spain population figures Overveiw details 

In this section we highlight a couple of trends observed in both Total and foreign population in Spain from 2004 until 2025 period.

```{r Population figures overview, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Total population figures overview

Max_total_population <- INE_calc_fields  %>% select (Year,total_population) %>% filter(Year == max(INE_calc_fields$Year))
Max_total_population


Min_total_population <- INE_calc_fields  %>% select (Year,total_population) %>% filter(Year == min(INE_calc_fields$Year))
Min_total_population

# Foreign population figures overview
foreign_pop_first_year <- INE_calc_fields  %>% select (Year,foreign_population) %>% filter(Year == 2005)

foreign_pop_last_year <- INE_calc_fields  %>% select (Year,foreign_population) %>% filter(Year == 2025)
```

- As of 1st of January **`r (Max_total_population$Year)`**., the year for the latest available population figures, Total population in Spain was  **`r prettyNum(Max_total_population$total_population, big.mark=",")`**. Up by 458,253 from previous year.

- In contrast, on the year **`r (Min_total_population$Year)`**, the first year on this series, total population in Spain was **`r prettyNum(Min_total_population$total_population, big.mark=",")`**. 

- In terms of foreign population, on the following year **`r (foreign_pop_first_year$Year)`**.there was a foreign population of **`r prettyNum(foreign_pop_first_year$foreign_population, big.mark=",")`** 

- At the end of the series on 1st January **`r (foreign_pop_last_year$Year)`**, latest foreign population figures in Spain was **`r prettyNum(foreign_pop_last_year$foreign_population, big.mark=",")`**. We will describe these population changes in absolute figures and percent change in the next section below.

### 3.1 Share of foreign population over total population in Spain

Next, we describe the foreign population percent share of the total population in Spain for the 2004-2025 period.


## 4. Year on year increase in total and forein population in Spain

In this section we will include several calculations to describe the year on year change in Total, national and foreign population in Spain. And also we will include population change for the entire period considered 2004-2025.

- This is a population increase of () from the initial figure of **`r prettyNum(Min_total_population$total_population, big.mark=",")`**, in **`r (Min_total_population$Year)`**.


