---
title: "Parameterized RMarkdown"
author: "PLR"
date: "2025-07-25"
output: html_document
params:
  data: mpg # options: mpg or diamonds
  toggle: TRUE # options: TRUE or FALSE
  year: 2020  # Introduced new parameter for year
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# The `r params$data` Data set was chose in `r params$year`

We use a parameter to specify the **`r params$data`** data set we are currently using. From this **`r params$data`** data we choose to plot count of *manufacturers* as a bar plot in ggplot as shown below:

```{r barplot, include = params$toggle}
mpg %>% 
  ggplot() +
  aes(x = manufacturer) +
  geom_bar( fill = "#0c4c8a") +
  theme_minimal()

```

- In the same way we can use the *toggle* parameter to enable or disbale plots, in this instance we are **disabling** the rendering of this heat map below, based on the diamonds dataset

```{r heatmap, include=!params$toggle}
diamonds %>% 
  ggplot() +
  aes(x = cut, y = color, fill = clarity) +
  geom_tile(size = 1L) +
  scale_fill_hue() +
  theme_minimal()

```

### Ideas

